Dissection: verified and terminating

### 2018-05-18 paper/main.tex
One last reread of intro.

Unfortunately, all in the garden is not rosy. The operator \ensuremath{\AF{\_+\_}}
needs both of its parameters to be fully evaluated before it can
reduce further. When evaluating large expressions, the size of the
stack used during execution grows linearly with the size of the input,
potentially leading to a stack overflow, causing the execution of the
program to halt unexpectedly.

To address this problem, we can rewrite the evaluator to be
\emph{tail recursive}. Modern compilers typically map tail recursive
functions to machine code that runs in constant memory. To write such
a tail recursive function, we need to introduce an explicit stack
storing both intemediate results and the subtrees that have not yet
been evaluated. We can define such a stack as follows:

@@

We can define a tail-recursive evaluation function by means of a
pair of mutually recursive functions \ensuremath{\AF{load}} and \ensuremath{\AF{unload}}. The \ensuremath{\AF{load}}
function traverses the input expression, pushing subtrees on the
stack; the \ensuremath{\AF{unload}} function unloads the stack, while accumulating a
(partial) result.

@@

Implementing this tail recursive evaluator has come at a price: Agda's
termination checker flags the \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} functions as
potentially non-terminating by highlighting them
\nonterm{orange}. Indeed, in the very last clause of the \ensuremath{\AF{unload}}
function a recursive call is made to arguments that are not
syntactically smaller. Furthermore, it is not clear at all that the
tail recursive evaluator produces the same result as our original
one. It is precisely these questions that this paper tackles by
means of the following novel contributions:

\begin{itemize}
\item We give a verified proof of termination of \ensuremath{\AF{tail\text{-}rec\text{-}eval}} using
  a carefully chosen \emph{well-founded relation}
  (Section~\ref{sec:basics}). After redefining \ensuremath{\AF{tail\text{-}rec\text{-}eval}} using
  this relation, we prove that it agrees with the \ensuremath{\AF{eval}} function on
  all inputs.
\item We generalize this construction, inspired by McBride's
  generalization of Danvy's work on abstract machines. To do so, we
  define a universe of algebraic data types and a generic fold
  (Section~\ref{sec:universe}). Subsequently we show how to turn any
  structurally recursive function defined using a fold into a tail
  recursive variant (Section~\ref{sec:dissection}). 
\item Finally, we sketch how the proofs of termination and semantics
  preservation from our example are generalized to the generic fold
  over arbitrary types in our universe (Section~\ref{correctness}).
\end{itemize}

All the constructions and proofs presented in this paper have been
implemented in Agda\todo{citation}. The corresponding code is freely
available online.\footnote{\todo{url}}

>>>>

Unfortunately, not all in the garden is rosy. The operator \ensuremath{\AF{\_+\_}}
needs both of its parameters to be fully evaluated before it can
reduce further. When evaluating large expressions, the size of the
stack used during execution grows linearly with the size of the input,
potentially leading to a stack overflow, causing the execution of the
program to halt unexpectedly.

To address this problem, we can manually rewrite the evaluator to be
\emph{tail recursive}. Modern compilers typically map tail recursive
functions to machine code that runs in constant memory. To write such
a tail recursive function, we need to introduce an explicit stack
storing both intemediate results and the subtrees that have not yet
been evaluated. We can define such a stack as follows:

@@

We can define a tail-recursive evaluation function by means of a
pair of mutually recursive functions, \ensuremath{\AF{load}} and \ensuremath{\AF{unload}}. The \ensuremath{\AF{load}}
function traverses the expressions, pushing subtrees on the
stack; the \ensuremath{\AF{unload}} function unloads the stack, while accumulating a
(partial) result.

@@

Implementing this tail recursive evaluator has come at a price: Agda's
termination checker flags the \ensuremath{\AF{load}} and \ensuremath{\AF{unload}} functions as
potentially non-terminating by highlighting them
\nonterm{orange}. Indeed, in the very last clause of the \ensuremath{\AF{unload}}
function a recursive call is made to arguments that are not
syntactically smaller. Furthermore, it is not clear at all that the
tail recursive evaluator produces the same result as our original
one. Writing such tail recursive functions by hand is both tedious and
error-prone. It is precisely these issues that this paper tackles
by making the following novel contributions:

\begin{itemize}
\item We give a verified proof of termination of \ensuremath{\AF{tail\text{-}rec\text{-}eval}} using
  a carefully chosen \emph{well-founded relation}
  (Section~\ref{sec:basics}). After redefining \ensuremath{\AF{tail\text{-}rec\text{-}eval}} using
  this relation, we prove that it agrees with the \ensuremath{\AF{eval}} function on
  all inputs.
\item We generalize this construction, inspired by McBride and Danvy's
  work on calculating abstract machines from their recursive
  counterparts. To do so, we define a universe of algebraic data types
  and a generic fold operation
  (Section~\ref{sec:universe}). Subsequently we show how to turn any
  structurally recursive function defined using a fold into its tail
  recursive counterpart (Section~\ref{sec:dissection}).
\item Finally, we sketch how the proofs of termination and semantics
  preservation from our example are generalized to the generic fold
  over arbitrary types in our universe
  (Section~\ref{correctness}).

\end{itemize}

All the constructions and proofs presented in this paper have been
implemented in and checked by Agda. The corresponding code is freely
available online.\footnote{\todo{url}}
